---
title: "Assignment 2"
author: "Isabel"
date: "9/19/2019"
output: html_document
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

First, load the `gapminder` and `tidyverse` packages. The `dplyr` package will be loaded via the `tidyverse` package. 

```{r load packages, warning = FALSE, message = FALSE}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(DT))
# load your packages here:

```

Let's first have an overview of the data:

```{r}
datatable(as_tibble(gapminder))
```

Let's now focus on the following three countries: Singapore, Malaysia and Indonesia:

```{r}
filtered<-gapminder%>%
  filter(year>1969 & year<1980)%>%
  filter(country %in% c('Singapore', 'Malaysia', 'Indonesia'))
  
print(filtered)
  
```

We now only want the columns 'country' and 'gdpPercap' from the above dataset: 

```{r}

filtered%>%
  select(country, gdpPercap)

```

We want to see which countries have experienced a drop in life expectancy between 2007 and 1952: 

```{r}
gapminder%>%
  mutate(difference=lifeExp-lag(lifeExp, 1))%>%
  filter(country if difference<0)

```

