---
title: "Assignment 4"
author: "Isabel"
date: "10/2/2019"
output: html_document
---


```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(gapminder)
library(ggplot2)
library(DT)
```

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

## Exercise 1: Univariate Data Reshaping 

We want to first take a look at the data. 

```{r}
datatable(gapminder)
```

We now create a tibble with each row representing each year, and columns representing the life expectancies of different countries. 

```{r}
gapminder_wide <- gapminder %>% 
  pivot_wider(id_cols = c(year), 
              names_from = country, 
              values_from = lifeExp)%>%
  as_tibble()

gapminder_wide%>%
  datatable()
```

We now want to plot the life expectancy of Afghanistan against the life expectancy of Albania, both across time, using a scatterplot. 

```{r}
gapminder_wide%>%
  ggplot()+
  geom_point(aes(year, Afghanistan, color="Afghanistan"))+
  geom_point(aes(year, Albania, color="Albania"))+
  ylab("Life Expectancy")+
  ggtitle("Life Expectancy of Afghanistan and Albania plotted across time")+
    scale_color_manual("",
                       breaks=c("Afghanistan", "Albania"),
                       values=c("red", "blue"))

#I wanted to construct a legend without melting the dataset. I took reference from the following website: https://stackoverflow.com/questions/10349206/add-legend-to-ggplot2-line-plot
```

Let's now lengthen the data again. 

```{r}

gapminder_wide%>%
  pivot_longer(cols= "Afghanistan":"Zimbabwe",
               names_to="country",
               values_to="LifeExp")%>%
  as_tibble()%>%
  datatable()

```

## Exercise 2: Multivariate Data Reshaping 

We first make a tibble with each row representing one year, and columns for life expectancy and GDP per capita for Boliva, China and Switzerland. 

```{r}
gapminder_wider <- gapminder %>%
  filter(country==c("Bolivia", "China", "Switzerland"))%>%
  pivot_wider(id_cols = c(year), 
              names_from = country,
              names_sep = "_",
              values_from = c(lifeExp, gdpPercap))%>%
  arrange(year)%>%
                as_tibble()

gapminder_wider%>%
  datatable()

```

We then re-lengthen the data.

```{r}
gapminder_wider%>%
  pivot_longer(cols = -year, 
               names_to=c(".value", "country"),
               names_sep="_")%>%
  datatable()

```

## Exercise 3: Table Joins

```{r}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

In order to add a new column for email address for each row of guest in the guest list, I first need to modify my email tibble such that every row only has one guest name. Then I want to rename the column 'guest' to 'name' so that I can join the two tibbles. 

```{r}

email_indvl<-email%>%
  separate_rows(guest, sep = ", ")%>%
  rename(name=guest)%>%
  as_tibble()

guest%>%
  left_join(email_indvl, by="name")%>%
  as_tibble()%>%
  datatable()
```

I use an anti-join to find out who we have mails for, but are not on the guestlist.

```{r}
email_indvl%>%
  anti_join(guest, by="name")%>%
  as_tibble()
```

I then create a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).

```{r}
guest%>%
  full_join(email_indvl, by="name")%>%
  select(-email)%>%
  as_tibble()%>%
  datatable()
```


