---
title: "Assignment 4"
author: "Isabel"
date: "10/2/2019"
output: html_document
---


```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(gapminder)
library(ggplot2)
```

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

## Exercise 1: Univariate Data Reshaping 

We want to first take a look at the data. 

```{r}
gapminder
```

We now create a tibble with each row representing each year, and columns representing the life expectancies of different countries. 

```{r}
gapminder_wide <- gapminder %>% 
  pivot_wider(id_cols = c(year), 
              names_from = country, 
              values_from = lifeExp)%>%
  as_tibble()

gapminder_wide
```

We now want to plot the life expectancy of Afghanistan against the life expectancy of Albania, both across time, using a scatterplot. 

```{r}
gapminder_wide%>%
  ggplot()+
  geom_point(aes(year, Afghanistan, color="blue"))+
  geom_point(aes(year, Albania, color="red"))+
  ylab("Life Expectancy")

#I want to put a legend but I don't know how!
```

Let's now lengthen the data again. 

```{r}

gapminder_wide%>%
  pivot_longer(cols= Afghanistan:Zimbabwe,
               names_to="country",
               values_to="LifeExp")

```

## Exercise 2: Multivariate Data Reshaping 

We first make a tibble with each row representing one year, and columns for life expectancy and GDP per capita for Boliva, China and Switzerland. 

```{r}
gapminder_wider <- gapminder %>%
  filter(country==c("Bolivia", "China", "Switzerland"))%>%
  pivot_wider(id_cols = c(year), 
              names_from = country,
              names_sep = "_",
              values_from = c(lifeExp, gdpPercap))%>%
  arrange(year)%>%
                as_tibble()

gapminder_wider

```

We then re-lengthen the data.

```{r}
gapminder_wider%>%
  pivot_longer(cols = -year, 
               names_to=c(".value", "country"),
               names_sep="_")

```

## Exercise 3: Table Joins


